Energy Clustering

### 2017-10-26 energy_clustering/draft_v2/energy_clustering.tex
draft v2

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen. A well-known kernel based clustering method
is kernel $k$-means, which is preciselly $k$-means 
formulated in the feature space \cite{Girolami}. 
Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd}
of grouping points that are closer to a cluster center.
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics, in Euclidean spaces,
was considered in \cite{Kgroups}, 
the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. This immediatelly
establishes the connection between energy statistics based clustering,
or \emph{energy clustering} for short, with kernel methods. Moreover,
our formulation holds for general semimetric spaces of negative type.
We also show that such QCQP is equivalent to kernel $k$-means 
optimization problem, however, the kernel is fixed by 
energy statistics.
The equivalence between kernel $k$-means, spectral clustering, and graph 
partitioning problems is well-known \cite{Dhillon,Dhillon2}. We further
demonstrate 
how these relations arise from a weighted version of energy statistics.

@@

% @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.  i'm not sure the best way to visualize it, maybe play with a few options.
% @jovo: I didn't do this yet and actually not sure how to do it.
% Are you thinking about this: 
% https://en.wikipedia.org/wiki/Kernel_density_estimation
% Also, I'm not sure if the kernel from energy integrates to 1, etc. and is
% this really worthy it?
% @gui: yes, it is really worth it, that is why i suggested it. it is trivial to do.  currently, the reader has no idea *why* the different algs get different answers.

% @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
% @jovo: yes, and GMM would do great! In 1D is very hard to find settings where
% GMM and k-means perform poorly, so we have to cherry pick a little. In 1D you
% basically do not have much choice, if data has minimal separation k-means and
% GMM do well. The purpose is to
% illustrate that there are settings where k-means/GMM do poorly and energy
% does not. If I choose the same parameters as before all algorithms are doing
% great, sometimes GMM can even do better than energy. All I did was to
% translate one of the lognormals mu2=-1.5 and give it a big variance, so both
% distributions overlap a lot, but the overlaping is essentially on the tail 
% of one of them
% @gui: ok, well why not use the same params from fig 1b in fig 1a then?

@@

We now consider two simple experiments where we sample $n$ points
from a two-class mixture.
We plot the average accuracy \eqref{eq:accuracy} versus $n$, 
with error bars indicating standard error.
The data is clustered using $\mathcal{E}^{1D}$-clustering algorithm,
GMM and $k$-means. 
For GMM and $k$-means we use the implementations
from the well-known \emph{scikit-learn} library in 
Python \cite{scikit-learn}, where $k$-means is initialized through
$k$-means++ procedure \cite{Vassilvitskii}, and GMM is initialized
with the output of $k$-means. We run both algorithms
$5$ times with different initializations and pick the answer with best 
objective function value.
Notice that $\mathcal{E}^{1D}$-clustering does not require random
initialization so we only run it once.
For each $n$ we use use $100$ Monte Carlo runs.
In  Fig.~\ref{fig:1d}a 
we have the results for data sampled from the Gaussian mixture
\begin{equation}
\label{eq:two_normal}
x \stackrel{iid}{\sim} 
\tfrac{1}{2}\mathcal{N}\big(\mu_1,\sigma_1^2\big) 
+\tfrac{1}{2}\mathcal{N}\big(\mu_2,\sigma_2^2\big), 
\qquad 
\mu_1 = 1.5, 
\sigma_1=0.3, 
\mu_2 = 0, 
\sigma_2 = 1.5.
\end{equation}
In this case the optimal accuracy 
obtained from Bayes classification error
is $ \approx 0.88$, indicated by the dashed line in the plot.
The three methods
perform closely, with a slight advantage of GMM, as expected since
it is a consistent model to the data, and 
$\mathcal{E}^{1D}$-clustering performs slightly better
than $k$-means. On the other hand, in Fig.~\ref{fig:1d}b
we consider a mixture of lognormal distributions,
\begin{equation}
\label{eq:two_lognormal}
x \stackrel{iid}{\sim} 
\tfrac{1}{2} \exp\left\{\mathcal{N}\big( \mu_1,\sigma_1^2\big)\right\} 
+\tfrac{1}{2} \exp\left\{\mathcal{N}\big( \mu_2,\sigma_2^2\big)\right\}, 
\qquad 
\mu_1 = 1.5, 
\sigma_1=0.3, 
\mu_2 = 0, 
\sigma_2 = 1.5.
\end{equation}
The optimal Bayes accuracy is again $\approx 0.88$.
We can now see that $\mathcal{E}^{1D}$-clustering
is still  very accurate, while
GMM and $k$-means basically cluster at chance.
The two simple experiments of Fig.~\ref{fig:1d} illustrate
how energy clustering is more flexible compared to $k$-means and GMM.

\begin{figure}
\begin{minipage}{0.49\textwidth}
~~~~~\includegraphics[width=.95\textwidth]{hist_normal.pdf}\\[-1em]
\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
~~~~~\includegraphics[width=.95\textwidth]{hist_lognormal.pdf}\\[-1em]
\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-1em](b)
\end{minipage}
\caption{
\label{fig:1d}
$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM.
We plot the average 
accuracy \eqref{eq:accuracy}, over $100$ Monte Carlo trials, 
versus the number of sampled points. 
Error bars are standard error. The dashed line indicates 
Bayes accuracy ($\approx 0.88$ in both cases). (a) Clustering results for data normally distributed as in 
\eqref{eq:two_normal}, where we show the histograms of each component
above the experimental plot.
(b) Data lognormally distributed as in \eqref{eq:two_lognormal}
with histograms of the components in the mixture.
}
\end{figure}

@@

We briefly mention that we compared $\mathcal{E}^H$-clustering,
as described in Algorithm~\ref{algo}, to
$\mathcal{E}^{1D}$-clutering, described in Algorithm~\ref{algo1d}, 
for several univariate distributions.
Both perform very closely.
However, we omit these results since we will analyse more interesting 
scenarios in high dimensions.

From the results of \cite{Telgarsky}, summarized in the end of the previous
section, we expect the improvement of Hartigan's 
over Lloyd's method to be more accentuated in high dimensions.
Thus, we analyze
how the algorithms degrade as the number of dimensions increase while
keeping the number of points in each cluster fixed.
Consider data from the Gaussian mixture
\begin{equation}
\label{eq:gauss1}
\begin{split}
x  &\stackrel{iid}{\sim} 
\tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \\
\Sigma_1&=\Sigma_2 = I_D, \quad
\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top, \quad
\mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top.
\end{split}
\end{equation}
Note that the Bayes error
is fixed as $D$ increases, yielding an optimal 
accuracy of $\approx 0.86$.
We sample $200$ points on each trial.
The results are shown in Fig.~\ref{fig:gauss}a. We can see that
$\mathcal{E}^H$ and spectral clustering have practically
the same performance, which is 
higher than $\mathcal{E}^L$-clustering (kernel
$k$-means). Moreover, $\mathcal{E}^H$ outperforms 
$k$-means and GMM, where the improvement is noticeable specially in 
high dimensions.
Note that in this setting $k$-means and GMM are consistent models to the data,
however, energy clustering degrades much less as dimension increases.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{normal_highdim_mean.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{normal_highdim_cov.pdf}\\[-1.0em] (b)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of $\mathcal{E}^H$-clustering, $\mathcal{E}^L$-clustering (kernel
$k$-means), spectral clustering, $k$-means and GMM in
high dimensional Gaussian settings. We plot the mean accuracy versus
the number of dimensions, with error bars indicating standard 
error from $100$ Monte
Carlo runs.
(a) Data normally distributed as in \eqref{eq:gauss1}, with Bayes
accuracy $\approx 0.86$, over the range $D \in [10,200]$. 
(b) Data normally distributed
as in \eqref{eq:gauss2}, with Bayes accuracy $\approx 0.95$,
over the range $D \in [10, 700]$.
% @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".  i have no intuition how difficult these are.
% @gui: https://www.google.com/search?q=pairs+plots&rlz=1C5CHFA_enUS731US731&source=lnms&tbm=isch&sa=X&ved=0ahUKEwjwo9u5w-DWAhXMKyYKHYkhAKMQ_AUICigB&biw=1440&bih=776
% to show how well separated the classes are.
}
\end{figure}

Still for a two-class Gaussian mixture, 
we now allow the diagonal entries of one of the covariances to
have different values by choosing
\begin{equation}
\label{eq:gauss2}
\begin{split}
x & \stackrel{iid}{\sim} 
\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)+
\tfrac{1}{2}\mathcal{N}(\mu_2,\Sigma_2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = I_D, \quad
\Sigma_2 = \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
1.969, 4.035,   4.237,  2.813,  3.637),
\end{split}
\end{equation}
We simply chose a fixed set of $10$ numbers uniformly at random on 
the interval $[1,5]$ for
the diagonal of $\widetilde{\Sigma}_{10}$, and any other choice would give 
analogous results. We sample a total of $200$ points from \eqref{eq:gauss2}
on each trial.
The Bayes error is kept fixed when increasing $D$ yielding
an optimal accuracy $\approx 0.95$.
In Fig.~\ref{fig:gauss}b we see that GMM performs better in low dimensions, 
but it quickly degenerates as $D$ increases. 
The same is true for $k$-means and $\mathcal{E}^L$-clustering.
However, $\mathcal{E}^H$ and spectral clustering remains much
more stable in high dimensions.
Notice that a naive implementation of GMM should not be able to 
estimate the covariances when $D \gtrsim 100$, however, scikit-learn library
uses $k$-means output as initialization, therefore the
output of GMM in this implementation is at least as good as $k$-means and
the algorithm is more robust in high dimensions.

>>>>

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen. A well-known kernel based clustering method
is kernel $k$-means, which is precisely $k$-means 
formulated in the feature space \cite{Girolami}. 
Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd}
of grouping points that are closer to a cluster center.
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics, in Euclidean spaces,
was considered in \cite{Kgroups}, 
the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. This immediately
establishes the connection between energy statistics based clustering,
or \emph{energy clustering} for short, with kernel methods. Moreover,
our formulation holds for general semimetric spaces of negative type.
We also show that such QCQP is equivalent to kernel $k$-means 
optimization problem, however, the kernel is fixed by 
energy statistics.
The equivalence between kernel $k$-means, spectral clustering, and graph 
partitioning problems is well-known \cite{Dhillon,Dhillon2}. We further
demonstrate 
how these relations arise from a weighted version of energy statistics.

@@

@@

We now consider two simple experiments where we sample $n$ points
from a two-class mixture.
We plot the average accuracy \eqref{eq:accuracy} versus $n$, 
with error bars indicating standard error.
The data is clustered using $\mathcal{E}^{1D}$-clustering algorithm,
GMM and $k$-means. 
For GMM and $k$-means we use the implementations
from the well-known \emph{scikit-learn} library in 
Python \cite{scikit-learn}, where $k$-means is initialized through
$k$-means++ procedure \cite{Vassilvitskii}, and GMM is initialized
with the output of $k$-means. We run both algorithms
$5$ times with different initializations and pick the answer with best 
objective function value.
Notice that $\mathcal{E}^{1D}$-clustering does not require random
initialization so we only run it once.
For each $n$ we use use $100$ Monte Carlo runs.
In  Fig.~\ref{fig:1d}a 
we have the results for data sampled from the Gaussian mixture
\begin{equation}
\label{eq:two_normal}
x \stackrel{iid}{\sim} 
\tfrac{1}{2}\mathcal{N}\big(\mu_1,\sigma_1^2\big) 
+\tfrac{1}{2}\mathcal{N}\big(\mu_2,\sigma_2^2\big), 
\qquad 
\mu_1 = 1.5, 
\sigma_1=0.3, 
\mu_2 = 0, 
\sigma_2 = 1.5.
\end{equation}
In this case the optimal accuracy 
obtained from Bayes classification error
is $ \approx 0.88$, indicated by the dashed line in the plot.
The three methods
perform closely, with a slight advantage of GMM, as expected since
it is a consistent model to the data, and 
$\mathcal{E}^{1D}$-clustering performs slightly better
than $k$-means. 
In Fig.~\ref{fig:1d}c we show a density estimation from clustering
$1000$ points from this mixture using the three algorithms.
Notice that all of them are able to distinguish the two classes.
On the other hand, in Fig.~\ref{fig:1d}b
we consider a mixture of lognormal distributions,
\begin{equation}
\label{eq:two_lognormal}
x \stackrel{iid}{\sim} 
\tfrac{1}{2} \exp\left\{\mathcal{N}\big( \mu_1,\sigma_1^2\big)\right\} 
+\tfrac{1}{2} \exp\left\{\mathcal{N}\big( \mu_2,\sigma_2^2\big)\right\}, 
\qquad 
\mu_1 = 1.5, 
\sigma_1=0.3, 
\mu_2 = 0, 
\sigma_2 = 1.5.
\end{equation}
The optimal Bayes accuracy is again $\approx 0.88$.
We can now see that $\mathcal{E}^{1D}$-clustering
is still  very accurate, while
GMM and $k$-means basically cluster at chance.
Density estimation after clustering $1000$ points this
mixture using the three algorithms 
are
is shown in Fig.~\ref{fig:1d}d. Note that only $\mathcal{E}^{1D}$-clustering
was able to distinguish the two classes. $k$-means and GMM put most of the
points in a single cluster, and points on the tail of the second component
of \eqref{eq:two_lognormal} in the other cluster.
The experiments of Fig.~\ref{fig:1d} illustrate
how energy clustering is more flexible compared to $k$-means and GMM.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_normal.pdf}\\[-1em](a)\\
~~~~~\includegraphics[width=.9\textwidth]{normal_density.pdf}\\[-1.5em](c)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\[-1em](b)\\
~~~~~\includegraphics[width=.9\textwidth]{lognormal_density.pdf}\\[-1.5em](d)
\end{minipage}
\caption{
\label{fig:1d}
$\mathcal{E}^{1D}$-clustering versus $k$-means and GMM.
(a,b) We plot the mean accuracy
\eqref{eq:accuracy} over $100$ Monte Carlo trials, 
versus the number of sampled points. 
Error bars are standard error. The dashed line indicates 
Bayes accuracy ($\approx 0.88$ in both cases). (a) Clustering results for data normally distributed as in 
\eqref{eq:two_normal}.
(b) Data lognormally distributed as in \eqref{eq:two_lognormal}.
(c) Density estimation of each component in the mixture \eqref{eq:two_normal} 
after clustering $1000$ sampled points using the three algorithms, compared
to the ground truth. 
(d) The same but for lognormal data \eqref{eq:two_lognormal}.
}
\end{figure}

@@

We briefly mention that we compared $\mathcal{E}^H$-clustering,
as described in Algorithm~\ref{algo}, to
$\mathcal{E}^{1D}$-clustering, described in Algorithm~\ref{algo1d}, 
for several univariate distributions.
Both perform very closely.
However, we omit these results since we will analyse more interesting 
scenarios in high dimensions.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{pairsplot1.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{pairsplot2.pdf}\\[-1.0em] (a)
\end{minipage}
\caption{
\label{fig:pairsplot}
Pair plots for the first $5$ dimensions. (a) Data normally distributed
as in \eqref{eq:gauss1}. (b) Data normally distributed as in 
\eqref{eq:gauss2}. We sample $200$ points for both cases.
We can see that there is a considerable overlap between the clusters.
}
\end{figure}

From the results of \cite{Telgarsky}, summarized in the end of the previous
section, we expect the improvement of Hartigan's 
over Lloyd's method to be more accentuated in high dimensions.
Thus, we analyze
how the algorithms degrade as the number of dimensions increase while
keeping the number of points in each cluster fixed.
Consider data from the Gaussian mixture
\begin{equation}
\label{eq:gauss1}
\begin{split}
x  &\stackrel{iid}{\sim} 
\tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \\
\Sigma_1&=\Sigma_2 = I_D, \quad
\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top, \quad
\mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top.
\end{split}
\end{equation}
To get some intuition about how separated data points from each class
are, we show scatter plots between the first $5$ dimensions in 
Fig.~\ref{fig:pairsplot}a.
Note that the Bayes error
is fixed as $D$ increases, yielding an optimal 
accuracy of $\approx 0.86$.
We sample $200$ points on each trial.
The results are shown in Fig.~\ref{fig:gauss}a. We can see that
$\mathcal{E}^H$ and spectral clustering have practically
the same performance, which is 
higher than $\mathcal{E}^L$-clustering (kernel
$k$-means). Moreover, $\mathcal{E}^H$ outperforms 
$k$-means and GMM, where the improvement is noticeable specially in 
high dimensions.
Note that in this setting $k$-means and GMM are consistent models to the data,
however, energy clustering degrades much less as dimension increases.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{normal_highdim_mean.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{normal_highdim_cov.pdf}\\[-1.0em] (b)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of $\mathcal{E}^H$-clustering, $\mathcal{E}^L$-clustering (kernel
$k$-means), spectral clustering, $k$-means and GMM in
high dimensional Gaussian settings. We plot the mean accuracy versus
the number of dimensions, with error bars indicating standard 
error from $100$ Monte
Carlo runs.
(a) Data normally distributed as in \eqref{eq:gauss1}, with Bayes
accuracy $\approx 0.86$, over the range $D \in [10,200]$. 
(b) Data normally distributed
as in \eqref{eq:gauss2}, with Bayes accuracy $\approx 0.95$,
over the range $D \in [10, 700]$.
}
\end{figure}

Still for a two-class Gaussian mixture, 
we now allow the diagonal entries of one of the covariances to
have different values by choosing
\begin{equation}
\label{eq:gauss2}
\begin{split}
x & \stackrel{iid}{\sim} 
\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)+
\tfrac{1}{2}\mathcal{N}(\mu_2,\Sigma_2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = I_D, \quad
\Sigma_2 = \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
1.969, 4.035,   4.237,  2.813,  3.637).
\end{split}
\end{equation}
We simply chose a fixed set of $10$ numbers uniformly at random on 
the interval $[1,5]$ for
the diagonal of $\widetilde{\Sigma}_{10}$, and any other choice would give 
analogous results. 
We show pair plots of this data in Fig.~\ref{fig:pairsplot}b.
We sample a total of $200$ points from \eqref{eq:gauss2}
on each trial.
The Bayes error is kept fixed when increasing $D$ yielding
an optimal accuracy $\approx 0.95$.
In Fig.~\ref{fig:gauss}b we see that GMM performs better in low dimensions, 
but it quickly degenerates as $D$ increases. 
The same is true for $k$-means and $\mathcal{E}^L$-clustering.
However, $\mathcal{E}^H$ and spectral clustering remains much
more stable in high dimensions.
Notice that a naive implementation of GMM should not be able to 
estimate the covariances when $D \gtrsim 100$, however, scikit-learn library
uses $k$-means output as initialization, therefore the
output of GMM in this implementation is at least as good as $k$-means and
the algorithm is more robust in high dimensions.
